---
title: "Homework 3"
author: "Samina Rashiq"
output: github_document
---

## Problem 1

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)
```


```{r ny_noaa, include=FALSE}
data("ny_noaa")

# Calculate the number of unique weather stations
num_stations <- ny_noaa |>
  distinct(id) |>
  nrow()

# Calculate missing data percentages for relevant columns
missing_prcp <- ny_noaa |>
  summarize(missing = mean(is.na(prcp)) * 100) |>
  pull(missing)

missing_snow <- ny_noaa |>
  summarize(missing = mean(is.na(snow)) * 100) |>
  pull(missing)

# Correct column name 'snwd' for snow depth
missing_snwd <- ny_noaa |>
  summarize(missing = mean(is.na(snwd)) * 100) |>
  pull(missing)

missing_tmax <- ny_noaa |>
  summarize(missing = mean(is.na(tmax)) * 100) |>
  pull(missing)

missing_tmin <- ny_noaa |>
  summarize(missing = mean(is.na(tmin)) * 100) |>
  pull(missing)

# Get the total number of rows
total_rows <- nrow(ny_noaa)

```

This dataset is a subset of National Oceanic and Atmospheric Association (NOAA) weather data from January 1, 1981 to December 31, 2010. The data is from **`r num_stations` weather stations** in New York state and includes **`r format(total_rows, big.mark=",")` rows** and **7 columns**. 

Key variables include weather station ID (`id`), date of observation (`date`), precipitation in tenths of millimeters (`prcp`), snowfall in millimeters (`snow`), snow depth in millimeters (`sndw`), maximum temperature in tenths of degrees Celsius (`tmax`), and minimum temperature in tenths of degrees Celsius (`tmin`).

While each row contains a weather station ID and date, there is significant missing data across the remaining variables. Specifically, the columns for precipitation, snowfall, snow depth, maximum temperature, and minimum temperature are **`r round(missing_prcp, 2)`%**, **`r round(missing_snow, 2)`%**, **`r round(missing_snwd, 2)`%**, **`r round(missing_tmax, 2)`%**, and **`r round(missing_tmin, 2)`%** empty respectively. 

Missing data for precipitation, snowfall, and snow depth is likely due to days where those weather events did not occur and may not be a significant issue. However, the substantial amount of missing data for both maximum and minimum temperature poses a challenge, as nearly half of the rows lack this information.

```{r data_cleaning, echo=FALSE}
# Create new variables for year, month, and day
ny_noaa <- ny_noaa |>
  mutate(
    year = year(date),
    month = month(date),
    day = day(date),
    snow_mm = snow / 10          # Convert snowfall from tenths of mm to mm
  )

# Create a new variable for snow in millimeters
common_snowfall <- ny_noaa |>
  mutate(snow_mm = snow / 10) |>  # Convert snowfall to millimeters
  filter(!is.na(snow_mm)) |>      # Remove missing values
  count(snow_mm) |>               # Count occurrences of each snowfall value in mm
  arrange(desc(n)) |>             # Arrange in descending order
  rename("Snowfall (mm)" = snow_mm, "Frequency of Snowfall Value" = n)
```

#### Most commonly observed snowfall values in NY state
```{r snowfall_frequency_table, echo=FALSE}
# Display the table of the most common snowfall values
common_snowfall |> 
  head(10) |>  # Show the top 10 most common values
  knitr::kable()
```
0mm of snowfall is the most common value because snow generally only falls in certain months (November to March) and does not happen daily. This is followed by 2.5mm, 1.3mm, 5.1mm, and 7.6mm. These values are small and this makes sense because NY state is not very snowy compared to other places (e.g., Quebec, 10-30mm per day in these months). 

```{r find_max_temps, echo=FALSE}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Ensure 'tmax' is numeric throughout processing
ny_noaa <- ny_noaa |>
  mutate(tmax = as.numeric(tmax))

# Calculate average max temperatures for January for each station across years
avg_jan <- ny_noaa |>
  filter(month == 1) |>
  group_by(id) |>
  summarise(Avg_Max_Temp_Jan = mean(tmax / 10, na.rm = TRUE))  # Calculate mean, divide by 10

# Calculate average max temperatures for July for each station across years
avg_jul <- ny_noaa |>
  filter(month == 7) |>
  group_by(id) |>
  summarise(Avg_Max_Temp_Jul = mean(tmax / 10, na.rm = TRUE))  # Calculate mean, divide by 10

# Merge the two datasets on 'id' to get the average max temperatures for January and July
avg_max_temps <- inner_join(avg_jan, avg_jul, by = "id")

# Remove rows where there are missing maximum temperatures
avg_max_temps_clean <- avg_max_temps |>
  filter(!is.na(Avg_Max_Temp_Jan) & !is.na(Avg_Max_Temp_Jul))
```

#### Average max temperature in January and July for each station from 1981-2010
```{r histograms_max_temps, echo=FALSE}
# Calculate average max temperatures for January for each station across years
avg_jan <- ny_noaa |>
  filter(month == 1) |>
  group_by(id) |>
  summarise(Avg_Max_Temp_Jan = mean(as.numeric(tmax) / 10, na.rm = TRUE))  # Ensure tmax is numeric

# Histogram for January
ggplot(avg_max_temps_clean, aes(x = Avg_Max_Temp_Jan)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Average Max Temperatures in January", x = "Average Max Temperature (째C)", y = "Frequency") +
  theme_minimal()

# Histogram for July
ggplot(avg_max_temps_clean, aes(x = Avg_Max_Temp_Jul)) +
  geom_histogram(binwidth = 1, fill = "red", color = "black") +
  labs(title = "Distribution of Average Max Temperatures in July", x = "Average Max Temperature (째C)", y = "Frequency") +
  theme_minimal()

```

The structure of the plots for average max temperatures for both January and July appear to follow a relatively symmetrical and normal looking distribution. Neither plot appears to have obvious outliers.


```{r}
# Convert tmin and tmax to numeric to avoid errors
ny_noaa <- ny_noaa |>
  mutate(
    tmin = as.numeric(tmin),
    tmax = as.numeric(tmax)
  )

# Panel 1: tmin (X) vs tmax (Y) using Hexbin plot for density
ggplot(ny_noaa, aes(x = tmin / 10, y = tmax / 10)) +
  geom_hex(bins = 50) +  # Hexbin plot to show density
  scale_fill_viridis_c() +
  labs(title = "tmax vs tmin (Hexbin Density Plot)", x = "Minimum Temperature (째C)", y = "Maximum Temperature (째C)") +
  theme_minimal()

# Filter snowfall values greater than 0 and less than 100
snow_filtered <- ny_noaa |>
  filter(snow > 0 & snow < 100) |>
  mutate(snow_mm = snow / 10)  # Convert snow to mm

# Panel 2: Snowfall distribution by year (Boxplot)
ggplot(snow_filtered, aes(x = factor(year), y = snow_mm)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Distribution of Snowfall (0 < Snow < 100mm) by Year", x = "Year", y = "Snowfall (mm)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for clarity


```


## Problem 2

```{r import_tidy_merge}

#import demographics data
nhanes_covar <- read.csv("nhanes_covar.csv", header = FALSE, skip = 4)

colnames(nhanes_covar) <- nhanes_covar[1, ]
nhanes_covar <- nhanes_covar[-1, ]

#assign variable classes to demographics data
nhanes_covar$SEQN <- as.character(nhanes_covar$SEQN)

nhanes_covar$sex <- factor(
  nhanes_covar$sex,
  levels = c("1", "2"),
  labels = c("Male", "Female")
)

nhanes_covar$age <- as.numeric(nhanes_covar$age)

nhanes_covar$BMI <- as.numeric(nhanes_covar$BMI)

nhanes_covar$education <- factor(
  nhanes_covar$education,
  levels = c("1", "2", "3", "4"),
  labels = c("High School", "Bachelor's", "Master's", "PhD"),
  ordered = TRUE  # Specifies that it is ordinal data
)

#filter demographics data for missing values
nhanes_covar_filtered <- nhanes_covar %>%
  filter(age >= 21 & 
         !is.na(sex) & 
         !is.na(age) & 
         !is.na(BMI) & 
         !is.na(education))

#import accelerometer data
nhanes_accel <- read.csv("nhanes_accel.csv")

#assign variable classes to accelerometer data
nhanes_accel$SEQN <- as.character(nhanes_accel$SEQN)

nhanes_accel <- nhanes_accel %>%
  mutate(across(-SEQN, as.numeric))

#merge
nhanes_merged <- merge(nhanes_covar_filtered, nhanes_accel, by = "SEQN")

```

```{r education_table_graph}
library(knitr)
library(ggplot2)

#make sex/education table
sex_education_table <-
    table(nhanes_merged$education,
    nhanes_merged$sex)

kable(sex_education_table, caption = "Number of Men and Women in Each Education Category")

#make grouped bar plot of education grouped by sex
ggplot(nhanes_merged, aes(x = education, fill = sex)) +
  geom_bar(position = "dodge") +  # 'dodge' creates side-by-side bars for each gender
  labs(title = "Age Distribution by Education and Sex",
       x = "Education Category",
       y = "Count",
       fill = "Sex") +  # Legend label
  theme_minimal()
```
The number of men and women appear to be relatively balanced across all four education categories. While gender differences do not appear to be an issue, there seems to be a higher proportion of participants at the master's level, which could bias the results.

```{r total_activity_scatter}
nhanes_merged$total_activity <- rowSums(nhanes_merged[, grep("^min", colnames(nhanes_merged))])

ggplot(nhanes_merged, aes(x = age, y = total_activity, color = sex)) +
  geom_point() +  
  geom_smooth(method = "lm", se = FALSE) + 
  facet_wrap(~education) + 
  labs(title = "Total Activity vs Age by Education Level and Sex",
       x = "Age",
       y = "Total Activity",
       color = "Sex") +  
  theme_minimal()
```
Total activity declines with age across all education levels. For those with a high school education, males start with higher activity but drop below females later in life. In contrast, females maintain higher activity levels for those with bachelor's and master's degrees. The decline in activity appears to slow as education levels increase.

```{r 24_hour_time_courses}

#gather minute-level data into a long format
nhanes_long <- nhanes_merged %>%
  pivot_longer(cols = starts_with("min"), 
               names_to = "minute", 
               values_to = "activity") %>%
  mutate(minute = as.numeric(gsub("min", "", minute)))

#calculate average activity per minute by education and sex
average_activity <- nhanes_long %>%
  group_by(education, sex, minute) %>%
  summarise(avg_activity = mean(activity, na.rm = TRUE))

#plot activity time courses
ggplot(average_activity, aes(x = minute, y = avg_activity, color = sex)) +
  geom_line() +  # Original line plot
  geom_smooth(se = FALSE, color = "black", size = 0.5) +  # Smooth trend in black and thinner
  facet_wrap(~education) +
  labs(title = "24-Hour Activity Time Course by Education Level and Sex",
       x = "Minute of Day",
       y = "Average Activity",
       color = "Gender") +
  theme_minimal() + 
  scale_x_continuous(breaks = seq(0, 1440, by = 720),
                     labels = c("12 AM", "12 PM", "12 AM"))
```

The graph suggests that those with a high school diploma as their highest education level are the most active overall for males and females around mid-day Those with bachelors and masters degrees seem to have a bimodal activity pattern around mid-day, while those with a high school diploma see a steady decrease in activity from noon onwards. The graph seems to suggest that rising and sleeping times are relatively similar accross education levels, based on where the peaks begin and end. In all three groups, females appear to be more active than males. Overall, it seems like gender is a smaller factor in determining activity throughout the day than is level of education. 

